Introduction to Markov state modeling with the PyEMMA software [Article v\versionnumber]

### 2018-11-29 manuscript/manuscript.tex

It is apparent that there are many choices involved in MSM construction such as what features should be used and how many states should be chosen.
In 2013, the variational approach to conformational dynamics (VAC) was derived,
which enabled an objective comparison among different state decomposition choices for models built with the same Markovian lag time~\cite{noe-vac}.
More recently, the more general variational approach to Markov processes (VAMP) has been developed in order to facilitate the approximation and comparison of reversible models for basis sets that are continuous,
as opposed to discrete states~\cite{vamp-preprint}.
The VAMP can thus be used to perform model selection.
Specifically, we use the VAMP-2 score, which captures the kinetic variance explained by the model.
However, the MSM lag time cannot be optimized using VAMP,
and must be chosen using a separate validation as described above~\cite{husic2017note}.

@@

For all of these approaches,
dimensionality reduction is performed by projecting the (mean free) features $\tilde{\mathbf{x}}(t)$
onto the leading $d$ independent components $\mathbf{U}_d=\left[\mathbf{u}_1 \dots \mathbf{u}_d\right]$,
\begin{equation}
\mathbf{y}(t) = \mathbf{U}_d^\top \tilde{\mathbf{x}}(t),
\end{equation}
where, in practice, $d$ is chosen such that a specific fraction of kinetic variance $c_d$ is retained (e.g., \SI{95}{\percent}).

@@

The estimation of an MSM requires the dynamics between microstates to be Markovian.
However, in case of a poor dimension reduction and/or discretization or short trajectories,
we cannot anticipate this to be the case.
We illustrate this point in notebook~07.

An alternative, which is much less sensitive to poor discretization,
is to estimate a hidden Markov model (HMM)~\cite{hmm-baum-welch-alg,jhp-spectral-rate-theory,noe-proj-hid-msm,bhmm-preprint}.
HMMs are less sensitive to the discretization error as they sidestep the assumption of Markovian dynamics in the discretized space (illustrated in Fig.~\ref{fig:hmm-scheme}).
Instead, HMMs assume that there is an underlying (hidden) dynamic process that is Markovian
and gives rise to our observed data, e.g., the ($n$~states) discretized trajectories $s(t)$.
This is a powerful principle as we know that there is indeed an underlying process that is Markovian:
our molecular dynamics trajectories.

@@

An HMM estimation always yields a model with a small number of (hidden) states
in which each state is considered to be metastable and,
thus, the number of hidden states is a new hyper-parameter which needs to be chosen carefully (see notebook~07).
As the HMMs---like MSMs---approximate the full phase-space dynamics,
we can similarly compute the metastable kinetics, apply TPT, visualize the network, and obtain physical observables.

For an extensive discussion of details about HMM properties and the estimation algorithm in general, we suggest Ref.~\cite{hmm-tutorial}.
For its specific application to the discretization of MSMs using HMMs, we suggest Ref.~\cite{noe-proj-hid-msm}.
A generalized extension for estimating this type of low dimensional projection from the data is given in Ref.~\cite{wu2015projected}.

@@

Using PyEMMA, we can simulate the relaxation of an observable if we had prepared our molecular system in a nonequilibrium initial condition.
The experimental counterpart of such a prediction could be a temperature or pressure jump experiment or a stopped flow assay.
To illustrate such an experiment, we initialize our molecular ensemble as the metastable distribution of~$\mathcal{S}_1$
and follow the predicted fluorescence signal as it relaxes to equilibrium (Fig.~\ref{fig:msm-exp-obs}b).
We see that the predicted relaxation signal has a much larger amplitude for the nonequilibrium initialization,
making it more likely to be experimentally measurable.

@@

\subsection{Summary}

@@

\section{Funding Information}
%%%%%%%
% Authors should acknowledge funding sources here. Reference specific grants.
%%%%%%%
TH acknowledges financial support from Deutsche Forschungsgemeinschaft (SFB/TRR 186, Project A12).
FN and BEH acknowledge funding from European Commission (ERC CoG 772230 "ScaleCell").
FN acknowledges funding from Deutsche Forschungsgemeinschaft (SFB 1114, Projects A04 and C03, NO 825/2-2).
SO acknowledges a postdoctoral fellowship from the Alexander von Humboldt Foundation.

>>>>

It is apparent that there are many choices involved in MSM construction such as what features should be used and how many states should be chosen.
In 2013, the variational approach to conformational dynamics (VAC) was derived,
which enabled an objective comparison among different state decomposition choices for models built with the same Markovian lag time~\cite{noe-vac}.
More recently, the more general variational approach to Markov processes (VAMP) has been developed in order to facilitate the approximation and comparison of reversible models for basis sets that are continuous,
as opposed to discrete states~\cite{vamp-preprint}.
The VAMP can thus be used to perform model selection.
Specifically, we use the VAMP-2 score, which captures the kinetic variance explained by the model~\cite{kinetic-maps}.
However, the MSM lag time cannot be optimized using VAMP,
and must be chosen using a separate validation as described above~\cite{husic2017note}.

@@

For all of these approaches,
dimensionality reduction is performed by projecting the (mean free) features $\tilde{\mathbf{x}}(t)$
onto the leading $d$ independent components $\mathbf{U}_d=\left[\mathbf{u}_1 \dots \mathbf{u}_d\right]$,
\begin{equation}
\mathbf{y}(t) = \mathbf{U}_d^\top \tilde{\mathbf{x}}(t),
\end{equation}
where, in practice, $d$ is preferably chosen such that a specific fraction of kinetic variance $c_d$ is retained (e.g., \SI{95}{\percent}).

@@

The estimation of an MSM requires the dynamics between microstates to be Markovian.
However, in case of a poor dimension reduction and/or discretization or short trajectories,
we cannot anticipate this to be the case.

An alternative, which is much less sensitive to poor discretization,
is to estimate a hidden Markov model (HMM)~\cite{hmm-baum-welch-alg,jhp-spectral-rate-theory,noe-proj-hid-msm,bhmm-preprint}.
HMMs are less sensitive to the discretization error as they sidestep the assumption of Markovian dynamics in the discretized space (illustrated in Fig.~\ref{fig:hmm-scheme}).
Instead, HMMs assume that there is an underlying (hidden) dynamic process that is Markovian
and gives rise to our observed data, i.e., the ($n$~states) discretized trajectories $s(t)$.
This is a powerful principle as we know that there is indeed an underlying process that is Markovian:
our molecular dynamics trajectories.

@@

An HMM estimation always yields a model with a small number of (hidden) states
in which each state is considered to be metastable and,
thus, the number of hidden states is a new hyper-parameter which needs to be chosen carefully.
As the HMMs---like MSMs---approximate the full phase-space dynamics,
we can similarly compute the metastable kinetics, apply TPT, visualize the network, and obtain physical observables.

For an extensive discussion of details about HMM properties and the estimation algorithm in general, we suggest Ref.~\cite{hmm-tutorial}.
For its specific application to the discretization of MSMs using HMMs, we suggest Ref.~\cite{noe-proj-hid-msm}.
A generalized extension for estimating this type of low dimensional projection from the data is given in Ref.~\cite{wu2015projected}. One of our tutorial notebooks, to be discussed in the next section, provides an example of HMM analysis.

@@

Using PyEMMA, we can simulate the relaxation of an observable from a nonequilibrium initial condition.
The experimental counterpart of such a prediction could be a temperature or pressure jump experiment or a stopped flow assay.
To illustrate such an experiment, we initialize our molecular ensemble as the metastable distribution of~$\mathcal{S}_1$
and follow the predicted fluorescence signal as it relaxes to equilibrium (Fig.~\ref{fig:msm-exp-obs}b).
We see that the predicted relaxation signal has a much larger amplitude for the nonequilibrium initialization,
making it more likely to be experimentally measurable.

@@

\subsection{Summary of the showcase notebook}

@@

\section{Funding Information}
%%%%%%%
% Authors should acknowledge funding sources here. Reference specific grants.
%%%%%%%
MKS acknowledges financial support from European Commission (ERC StG 307494 "pcCell").
TH acknowledges financial support from Deutsche Forschungsgemeinschaft (SFB/TRR 186, Project A12).
SO acknowledges a postdoctoral fellowship from the Alexander von Humboldt Foundation.
FN and MKS acknowledge funding from Deutsche Forschungsgemeinschaft (SFB 1114, Projects A04 and C03, NO 825/2-2).
FN and BEH acknowledge funding from European Commission (ERC CoG 772230 "ScaleCell").
